# Sentiment Analysis is Sick Yo


## Basic idea

A common and intuitive approach to text is <span class="emph">sentiment analysis</span>.  In a grand sense we are interested in the emotional content of some text, e.g. posts on Facebook, tweets, or movie reviews.  Most of the time, this is obvious when one reads it, but if you have hundreds of thousands or millions of strings to analyze, you'd like to be able to do so efficiently.

We will use the <span class="pack">tidytext</span> package for our demonstration.  It comes with a lexicon of positive and negative words that is actually a combination of multiple sources, one of which provides numeric ratings, while the others suggest different classes of sentiment.


```{r lexicon, echo=-1}
set.seed(1234)
library(tidytext)
sentiments %>% slice(sample(1:nrow(sentiments)))
```

The gist is that we are dealing with a specific, pre-defined vocabulary.  Of course, any analysis will only be as good as the lexicon. The goal is usually to assign a sentiment score to a text, possibly an overall score, or a generally positive or negative grade. Given that, other analyses may be implemented to predict sentiment via standard regression tools or machine learning approaches.

## Issues

### Context, sarcasm, etc.

Now consider the following.

```{r sent_is_sick}
sentiments %>% filter(word=='sick') 
```

Despite the above assigned sentiments, the word *sick* has been used at least since 1960s surfing culture as slang for positive.  A basic approach to sentiment analysis as described here will not be able to detect slang or other context like sarcasm.  However, with lots of training data for a particular context may allow one to correctly predict such sentiment.  In addition, there are, for example, slang lexicons, or one can simply add their own complements to any available lexicon.

### Lexicons

In addition, the lexicons are going to maybe be applicable to *general* usage of English in the western world.  Some might wonder where exactly these came from or who decided that the word *abacus* should be affiliated with 'trust'. You may start your path by typing `?sentiments` at the console if you have the <span class="pack">tidytext</span> package loaded.

## Example


### The first thing the baby did wrong

We demonstrate sentiment analysis with the text *The first thing the baby did wrong*, which is a very popular brief guide to parenting written by world renown psychologist [Donald Barthelme][Donald Barthelme] who, in his spare time, also wrote postmodern literature.  This particular text talks about an issue with the baby, whose name is Born Dancin', who likes to tear pages out of books. Attempts are made by her parents to rectify the situation, without much success, but things are finally resolved at the end.  The ultimate goal will be to see how sentiment in the text evolves over time.

How do we start? Let's look at the <span class="objclass">sentiments</span> data set in the <span class="pack">tidytext</span> package.


```{r inspect_sentiments}
library(tidytext)
sentiments %>% slice(sample(1:nrow(sentiments)))
```

The <span class="objclass">bing</span> lexicon is just *positive* or *negative*. The AFINN is numerical, with ratings -5:5 that are in the <span class="objclass">score</span> column. The others get more imaginative, but also more problematic. Why *assimilate* is *superfluous* is beyond me. It clearly should be negative given the [Borg](https://en.wikipedia.org/wiki/Borg_%28Star_Trek%29) connotations.

```{r superfluous}
sentiments %>% 
  filter(sentiment=='superfluous')
```

But I digress.  We start with the raw text, reading it in line by line.  In what follows we read in all the texts (three) in a given directory, such that each element of 'text' is the work itself, i.e. `text` is a list column[^text]. The <span class="func">unnest</span> function will more or less unravel the work to paragraph form.

```{r baby_sentiment_importraw, echo=T}
library(tidytext)
barth0 = 
  data_frame(file = dir('data/texts_raw/barthelme/', full.names = TRUE)) %>%
  mutate(text = map(file, read_lines)) %>%
  transmute(work = basename(file), text) %>%
  unnest(text) 
```

One of the things stressed in this document is the iterative nature of text analysis.  You will consistently take two steps forward, and then one or two back as you find issues that need to be addressed. For example in a subsequent step I found there were encoding issues[^encoding], so the following attempts to fix them.  In addition we want to <span class="emph">tokenize</span> the documents such that our <span class="emph">tokens</span> are sentences (e.g. as opposed to words or paragraphs). The reason for this is that I will be summarizing the sentiment at sentence level.


```{r barth_fix_encoding}
# Fix encoding, convert to sentences
barth = barth0 %>% 
  mutate(text = sapply(text, stringi::stri_enc_toutf8, is_unknown_8bit=TRUE, validate=T)) %>% 
  unnest_tokens(sentence, text, token='sentences')
```

The next step is to drill down to just the document we want, and tokenize to the word level.  However, I create a sentence id so that we can group on it later.

```{r get_the_baby}
# get baby doc, convert to words
baby = barth %>% 
  filter(work=='baby.txt') %>% 
  mutate(sentence_id = 1:n()) %>% 
  unnest_tokens(word, sentence, drop=F) %>% 
  ungroup 
```

Now that the data has been prepped, getting the sentiments is ridiculously easy.  But that is how it is with text analysis.  All the hard work is spent with the data processing.  Here all we need is an <span class="emph">inner join</span> of our words with a sentiment lexicon of choice. This process will only retain words that are also in the lexicon.  I use the numeric-based lexicon here. At that point we get a sum score of sentiment by sentence.


```{r baby_sentiment}
# get sentiment via inner join
baby_sentiment = baby %>% 
  inner_join(get_sentiments("afinn")) %>% 
  group_by(sentence_id, sentence) %>% 
  summarise(score=sum(score)) %>% 
  ungroup

```

The following plots sentiment over sentence (note that not every sentence will receive a sentiment score). You can read the sentence by hovering over the dot.

```{r plot_sentiment, echo=FALSE}
# plot sentiment over sentences
baby_sentiment %>%
  plot_ly(width='50%') %>% 
  add_paths(x=~sentence_id, y=~score,
            color=I('#00aaff')) %>%
  add_markers(x=~sentence_id, y=~score, 
              color=I('#ff5500'),
              size=I(15),
              hoverinfo=~ 'text', 
              text=~str_wrap(sentence),
              showlegend=F) %>% 
  theme_plotly()
```

In general the sentiment starts out negative as the problem is explained. It bounces back and forth a bit but ends on a positive note.  You'll see that some sentences' context are not captured.  For example, sentence 16 is 'But it didn't do any good'.  However *good* is going to be marked as a positive sentiment in any lexicon by default. In addition, the token length will matter.  Longer sentences are more likely to have some sentiment, for example.



## Exercises

### Romeo & Juliet

For this exercise I'll invite you to more or less follow along, as there is notable pre-processing that must be done.  We'll look at sentiment in Shakespeare's Romeo and Juliet.  I have a cleaner version in the raw texts folder, but we can take the opportunity to use the <span class="pack">gutenbergr</span> package to download it directly from Project Gutenberg, a storehouse for works that have entered the public domain.

```{r rnj_load, echo=-c(3,5)}
library(gutenbergr)
gw0 = gutenberg_works(title == "Romeo and Juliet")  # look for something with this title
gw0[,1:4]
rnj = gutenberg_download(gw0$gutenberg_id)
DT::datatable(rnj, 
              rownames=F, 
              options=list(dom='tp',
                           autoWidth = TRUE,
                           columnDefs = list(list(width = '50px', targets = 0))))
```

<br> 

We've got the text now, but there is still work to be done.  The following is a quick and dirty approach, but see the [Shakespeare section][Shakespeare Start to Finish] to see a more deliberate one.

We first slice off the initial parts we don't want like title, author etc. Then we get rid of other tidbits that would interfere, using a little regex as well to aid the process.

```{r rnj_clean, echo=-2}
rnj_filtered = rnj %>% 
  slice(-(1:49)) %>% 
  filter(!text==str_to_upper(text),            # will remove THE PROLOGUE etc.
         !text==str_to_title(text),            # will remove names/single word lines
         !str_detect(text, pattern='^(Scene|SCENE)|^(Act|ACT)|^\\[')) %>% 
  select(-gutenberg_id) %>% 
  unnest_tokens(sentence, input=text, token='sentences') %>% 
  mutate(sentenceID = 1:n())
DT::datatable(select(rnj_filtered, sentenceID, sentence), 
              rownames=F, 
              options=list(dom='tp',
                           autoWidth = TRUE,
                           columnDefs = list(list(width = '50px', targets = 0)))
)
```

<br> 


The following unnests the data to word tokens.  In addition, you can remove stopwords like a, an, the etc. However, some of the stopwords have sentiments, so you would get a bit of a different result if you retain them.  As Black Sheep once said, the choice is yours, and you can deal with this, or you can deal with that.


```{r rnj_stopwords}
# show some of the matches
stop_words$word[which(stop_words$word %in% sentiments$word)] %>% head(20)


# remember to call output 'word' or antijoin won't work without a 'by' argument
rnj_filtered = rnj_filtered %>% 
  unnest_tokens(output=word, input=sentence, token='words') %>%   
  anti_join(stop_words)
```

Now we add the sentiments via the <span class="func">inner_join</span> function.  Here I use 'bing', but you can use another, and you might get a different result.

```{r rnj_sentiment}
rnj_filtered %>% 
  count(word) %>% 
  arrange(desc(n))

rnj_sentiment = rnj_filtered %>% 
  inner_join(sentiments)
rnj_sentiment
```

```{r rnj_bing}
rnj_sentiment_bing = rnj_sentiment %>% 
  filter(lexicon=='bing')
table(rnj_sentiment_bing$sentiment)
```



Looks like this one is going to be a downer. The following visualizes (via <span class="pack">plotly</span>) the positive and negative sentiment scores as one progresses sentence by sentence through the work.

```{r rnj_sentiment_as_game}
rnj_sentiment_bing %>% 
  arrange(sentenceID) %>% 
  mutate(positivity = cumsum(sentiment=='positive'),
         negativity = cumsum(sentiment=='negative')) %>% 
  plot_ly() %>% 
  add_lines(x=~sentenceID, y=~positivity, name='positive') %>% 
  add_lines(x=~sentenceID, y=~negativity, name='negative') %>% 
  layout(yaxis = list(title='sentiment')) %>% 
  theme_plotly()
```

<br>

In general it's a close game until perhaps the midway point, when negativity takes over and despair sets in with the story.  By the end [[:SPOILER ALERT:]] Sean Bean is beheaded, Darth Vader reveals himself to be Luke's father, and Number One is *you*.

Here is the same information expressed as a difference.

```{r rnj_sentiment_diff, echo=F}
rnj_sentiment_bing %>% 
  arrange(sentenceID) %>% 
  mutate(positivity = cumsum(sentiment=='positive'),
         negativity = cumsum(sentiment=='negative')) %>% 
  plot_ly() %>% 
  add_lines(x=~sentenceID, y=~positivity-negativity) %>% 
  theme_plotly()
```

<br>

## Summary

In general, sentiment analysis can be a useful exploration of data, but it is highly dependent on the context and tools used.  Note also that 'sentiment' can be anything, it doesn't have to be positive vs. negative.  Any vocabulary may be applied, and so it has more utility than the usual implementation. 


[^text]: Don't name your column 'text' in practice. It is a base function in R, and the tidyverse will have problems with distinguishing the function from the column name.  I only do so for pedagogical reasons.

[^encoding]: There are almost always encoding issues in my experience.